---
title: "Exploring Categorical Variables Intro"
output: github_document
---
#Always start by running the following code to load the libraries. 

```{r libraries, include=FALSE}
library(tidyverse)
library(skimr)
library(janitor)
library(knitr)
library(kableExtra)
library(gridExtra)
library(openintro)
library(readr)
```

# As always, before diving into any questions, make sure you take a look at the dataset. 

This is a dataset that was generated by a group during the first part of the school year. 

```{r the data}
urlfile="https://raw.githubusercontent.com/jwadams25/high_school_Rstats/master/Copy%20of%20How%20stressed%20are%20Rivers%20Juniors_%20-%20Sheet1.csv"

stress_juniors <- read.csv(url(urlfile))

head(stress_juniors)

```

### Questions to Answer
##### 1. What are the observational units?
##### 2. What variables are included?
##### 3. What type of variables are they?  


# From a Dataset to a Table - One Categorical Variable

When exploring a question that involves one categorical variable, you'll first want to understand how many observations fall in each category and what proportion of them fall in each of those categories. Creating a table is a good first step you can take to do just that. 

The code below creates two tables, one for whether or not a student plays a sport and a second for showing whether or not a student is preparing for the ACTs or SATs. 

You'll notice, I gave each table a name (sport_count_prop and test_count_prop) and then wrote each of those names on their own lines so when you run the code, the table pops up. 

```{r tables for one categorical variable}
sport_count_prop <- stress_juniors %>%
                    count(sport) %>%
                    mutate(proportion = n / sum(n)) %>%
                    arrange(desc(proportion))
sport_count_prop

test_count_prop <- stress_juniors %>%
                    count(sat_act_prep) %>%
                    mutate(proportion = n / sum(n)) %>%
                    arrange(desc(proportion))
test_count_prop
```

### Question to Answer

#####  1. In the space where the code was written to generate those two tables, create a table showing the counts and proportions for those that play an instrument. 


# From a Dataset to a Visualization - One Categorical Variable

While a table can get you the numbers you are looking for and can be a useful visualization, you may also want to create a bar graph when looking at a categorical variable. The code below allows you to do that. 

```{r}

stress_juniors %>%
  ggplot(aes(x = sport)) +
  geom_bar(fill = "red") + 
  labs(x = "Do You Play a Sport", y = "Frequency")

stress_juniors %>%
  ggplot(aes(x = sat_act_prep)) +
  geom_bar(fill = "blue") + 
  labs(x = "Are You Preparing for the SAT and/or ACT", y = "Frequency")

```

### Questions to Answer

##### 1. What code changed the color of the bars? 

##### 2. This time, I have not given them names. Go into the code, give each a name and then write that name on a separate line as well so when you run the code again the bar graphs appear. Be sure the names you provide don't have spaces, are short, and help you remember what they represent. 

##### 3. In the space above where the code was written to make the two bar graphs, write new code to make a bar graph showing the number of people who do and do not play an instrument. 

# Visualizing Relationship Between 2 Categorical Variables 

After looking at each of the previous 2 categorical variables in isolation, the next logical thing to examine is whether there is a relationship between those two variables. Building a segmented bar graphs allows you to explore the relationship. 

```{r}
stress_juniors %>%
ggplot(aes(x = sport, fill = sat_act_prep)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c(COL["yellow","full"], COL["blue","full"])) + 
  labs(x = "Do you play a sport?", y = "Proportion", fill = "Are you prepping for the SATs or ACTs?")
 
```

### Questions to Answer
##### 1. What happens if you switch sport and sat_act_prep in the part of the code that says aes(x = sport, fill = sat_act_prep)?

##### 2. What happens if you change position = "fill" to position = "dodge"

##### 3. Which of the two graphs, fill or dodge, would be best in this situation and why?


# From a Dataset to a Table - Two Categorical Variables

## Creating a contigency table to expolore relationship between 2 categorial variables in greater detail. 

### Action Item
##### 1. Before going on, make sure the graph above is the segmented bar graph. 

With that graph now shown, you will want to be specific about the size of each section. In order do to that, you need to create a contigency table. 

```{r two way with conditional proportions}
stress_juniors %>%
  count(sport, sat_act_prep) %>% 
  group_by(sport) %>%
  mutate(proportion = n / sum(n)) %>% 
  select(-n) %>%
  pivot_wider(names_from = sport, values_from = proportion)
```

You may also want to look at the number of people that fall into each category. 

```{r two way with totals}

stress_juniors %>% 
  count(sport, sat_act_prep) %>% 
  pivot_wider(names_from = sport, values_from = n)

```


# Advanced Styling Options

## One Categorical Variable

For these demostrations we will use the data in the openintro packaged called immigration. This is the data used in the questions #2 done for homework in the following section of Introduction to Modern Statistics - https://openintro-ims.netlify.app/summarizing-visualizing-data.html#explore-categorical-data-exercises
Read here for more information about the dataset - http://openintrostat.github.io/openintro/reference/immigration.html


```{r immigration data}
head(immigration)
```

Let's start by seeing a breakdown of the number of people in each political affiliation in the sample of 910 registered voters in Tampa Florida. Like we did before, we'll make a table and a bar graph to do that. 

This time, however, we will arrange the table and graph from greatest to least (descending order). I'll also use the code coord_flip() to make the bars appear horizontally. These changes can help makes it much easier to draw conclusions.
 

```{r Bar plot ordering bars}
political_table <- immigration %>%
                    count(political) %>%
                    mutate(proportion = n / sum(n)) %>%
                    arrange(desc(proportion))


political_table

# When making this new bar graph, we'll want to reference the table we just created. Therefore, where you tyically write the name of the dataset you will write political_count because that is the name of our new dataset. 

political_bar_order <- immigration %>%
  ggplot(aes(x = fct_rev(fct_infreq(political)), fill = political)) +
  geom_bar() +
  theme_minimal() +
  coord_flip() +
  labs(x = "", 
       y = " Political Affiliation", 
       title = "Political Affiliations of Registered Votes in Tampa, Florida", 
       subtitle = "Based on a random sample of 910 registred voters in Tampa, Florida") +
  scale_fill_manual(values = c("red","navy", "gray")) +
  theme(legend.position = "none")

political_bar_order 


```
### Questions to Answer

##### 1. Reference the code above to make a table and bar graph for the response variable. Be sure the table lists proportions from greatest to least. For the bar graph, be sure each the bars are orered from least go greatest and all of other styling added to the graph above is included. *Hint: Be sure you add colors!

```{r create your own}

```


